\Large \bf One-Step, Three-Factor Authentication with Custom-Fit, In-Ear EEG

### 2017-02-17 paper.tex
Runaway argument? Can't compile PDF from tex file

\subsection*{Abstract}
We present a system that provides all three factors of authentication (knowledge, possession, and inherence) in a single step by using brain-based authentication via custom-fit, in-ear electroencephalography (EEG). We demonstrate its potential by collecting EEG data from seven participants using manufactured custom-fit earpieces with embedded electrodes. Across all participants, we achieve mean 0\% false acceptance and 0.017\% false rejection rates with data from only one earpiece with three electrodes. Furthermore, we find a 0\% false acceptance rate in a preliminary simulation of an impersonation attack. We also report some positive usability perspectives from our participants and discuss constructive directions for future work. Our results indicate that in-ear EEG could provide a discreet, convenient, and usable method for secure one-step, three-factor authentication.

@@

We find that we can achieve low false acceptance rates (FAR) and false rejection rates (FRR) with a single, 
three-electrode earpiece. Interestingly, we found that performance improves in the ear compared to a single electrode on the scalp, a site typically used for collecting EEG data.
Additionally, we find that a passthoughts could not be spoofed by imposters, even when the imposters had the user's earpiece and knew their chosen passthoughts.
We discuss future work to investigate the robustness of passthoughts in day-to-day life, and raise questions about how and why passthoughts work as well as they do.

@@

Following this, we investigate custom-fit in-ear EEG technology as the platform for investigating the feasibility, performance, and usability of one-step three-factor authentication worn unobtrusively. The system we propose and test here uses the choice of a mental task or "passthought" to perform as knowledge (factor one), the uniqueness of an individual's brain activity as measured by EEG as inherence (factor two), and the physical token of custom-fit earpieces, which could easily contain a hardware key-pair, as a possession factor (factor 3). Because three-factor authentication (3FA) requires the user to submit one distinct instance of each type of authenticator, it represents the strongest level of authentication security possible. The ability to utilize all three of these security factors in a single step by performing a mental task of a few seconds is a promising in pursuit of extremely strong security while maintaining a low amount of effort and obtrusiveness to the user.
 %% In addition to the dual knowledge and inherence factors established in previous work, this study incorporates a possession factor in the form of the EEG-sensing ear-piece(s) that are custom-fitted to and worn in their ear. These earpieces can serve as physical tokens in the same way as bank ATM cards and wearable hardware tokens by implementing a hardware key-pair.
 %% Furthermore, because the earpieces are custom-fitted to each individual, we predicted they would likely not be able to produce good electrical impedances when worn by a different individual.

@@

During the experiment, participants were seated in a comfortable position in a quiet room facing a laptop screen on which the instructions and stimuli were presented using PsychoPy \cite{peirce2007psychopy}. Each task was completed once in sets five trials each, and then each was completed again for another five trials. Each trial was 10 seconds in length, for a total of 10 trials and 100 seconds of data collected per task. The instructions were read aloud to the participant by the experimenter, and the experiment was advanced using a pointer held in the participant's lap to minimize motion artifacts in the data. The experimenter also recorded the participant's chosen secrets for the \textit{sport}, \textit{song}, \textit{face}, \textit{speech}, and \textit{sequence} tasks and reminded the participant of these for the second set of trials.

@@

In the previous section, we trained and tested a passthoughts classifier with each task as the passthought, for all participants. Our analyses revealed the left ear alone produced decent results. Focusing on the left ear, we filter our results for the best-performing tasks. We rank tasks by lowest FAR and, given a tie, the lowest FRR (Table \ref{tab:best}). Overall, we find  a mean FAR of zero and mean FRR of 0.017\%.

\begin{table}[h]
\begin{center}
\begin{tabular}{rrrl}
\textbf{P} & \textbf{FAR} & \textbf{FRR} & \textbf{Task}\\
\hline
1 & 0.000 & 0.000 & Listen\\
2 & 0.000 & 0.000 & Breathe\\
3 & 0.000 & 0.000 & Breathe\\
4 & 0.000 & 0.000 & Breathe\\
5 & 0.000 & 0.012 & Breathe\\
6 & 0.000 & 0.000 & Breathe\\
7 & 0.000 & 0.000 & Breathe\\
\end{tabular}
\end{center}
\caption{Best-performing passthought, with FAR and FRR, for each participant (P), using data from the left ear.}
\label{tab:best}
\end{table}

All best-performing tasks in our best-case set achieved perfect FAR and FRRs, with the exception of P5, whose best-performing task (\textit{breathe}) had a nonzero FRR.
\textit{Breathe} appeared as the best task across all participants, except P1. Given our training strategy, these results indicate that a given person's \textit{breathe} task is distinguishable not only among other tasks, but among \textit{breathe} tasks from other participants.
This task performed about as well as its open-eyes counterpart (Figure \ref{fig:meanByTask}), as did \textit{song - open} and \textit{sport} (though, interestingly, \textit{song} with eyes closed performed less well).

@@

Overall, our original training strategy achieves the lowest FAR (Table \ref{tab:compare}). Within-tasks FAR ten times higher, and within-participants FAR is one hundred times higher, though all scenarios still less than 1\%.
However, FRR is \textit{lower} in the within-tasks training strategy than in our original strategy's FRR. Within-participants again results in the highest FRR.

@@

While our left ear results establish that passthoughts achieve low FAR and FRR when tested against other participants' passthoughts, we do not know how robust passthoughts are against a spoofing attack, in which both a participant's custom-fit electrode, and details of that participant's chosen passthought, are leaked and used by an imposter.

@@

However, since this participant's data appeared in the initial pool, the classifier may have been trained on his or her recordings as negative examples. To explore the efficacy of an outsider spoofing recordings, we repeated the same protocol with an individual who did not appear in our initial set of participants (an "outside imposter,'' Table \ref{tab:imposter}). Again, find zero successful authentication attempts.

@@

In our analysis, some notable patterns emerged. First, \textit{breathe} tended to be the best-performing task among participants. Classifiers overall distinguished the \textit{breathe} task even compared to \textit{breathe} tasks from other participants, implying that the task is expressed differently for each participant, i.e. that this task has an inherence factor sufficient for authentication, even though the task does not have an explicit secondary knowledge factor as in \textit{song}, for example. Second, we were able to achieve good results by generating feature vectors based on only 500ms (300 voltage readings across the three electrodes). This short timespan is somewhat surprising, given that some tasks (like songs) presumably rely on changes or patterns over a longer period of time.

Furthermore, likely due to the use of conductive gel, low impedances are still achievable using other participants' custom-fit earpieces, despite the uniqueness of ear canal shapes between individuals \cite{Akkermans2005}.
Nevertheless, classifiers appear to resists spoofing attacks, indicating that task-related signals are unique to individuals.
In terms of establishing a possession factor, the custom-fit earpiecs could easily include a hardware keypair to sign authentication attempts in a three-factor, one-step scheme.

@@

One primary question surrounds how our passthoughts system performance will change with a greater number of users, and with more diverse data. Our system specifically trains on negative examples of incorrect users; we do not yet know how this approach will scale. At the same time, we must investigate the stability of EEG readings for a passthought are over time to establish long-term usability. We must also collect EEG data from the variety of different user states: ambulatory settings, during physical exertion or exercise, under the influence of caffeine or alcohol, etc.

% We provide cursory evidence that passthoughts are difficult to spoof, though further work should expand on ours with more subjects.
Another important question surrounds how passthoughts might be cracked.
Generally, we do not understand how an individual's passthought is drawn from the distribution of EEG signals that an individual produces throughout the day. 
Given a large enough corpus of EEG data, are some passthoughts as easy to guess as \textit{password1234} is for passwords?
Future work should perform statistical analysis on passthoughts, such as clustering (perhaps with t-SNE) to better understand the space of possible passthoughts.
This work will allow us simulate cracking attempts, and to develop empirically motivated strategies for prevention, e.g. locking users out after a certain number of attempts.
This work could also reveal interesting tradeoffs between the usability or accuracy of passthoughts and their security.

Finally, our work leaves room for some clear UX improvements.
Future work should try using dry electrodes, commonly found in consumer EEG devices, for comfort and usability. The electrodes could be grounded to the ear, instead of the forehead.
Additionally, future work could easily place a speaker inside our current custom-fit earbuds to produce working "hearables'' that can be used as ordinary headphones.

@@

As demonstrated by these preliminary results, custom-fit, in-ear EEG earpieces provide three factors of authentication in a single authentication single: thinking one's password, using the discreet form factor of an earpiece. In this paper, we demonstrate quite high authentication accuracy using a single sensing earpiece. By expanding our corpus of EEG readings (in population size, time, and diversity of settings), we hope to better understand the underlying distribution of EEG signals, so that we may better understand the security properties of passthoughts.

>>>>

\subsection*{Abstract}
We present a system that provides all three factors of authentication (knowledge, possession, and inherence) in a single step by using brain-based authentication via custom-fit, in-ear electroencephalography (EEG). We demonstrate its potential by collecting EEG data from seven participants using manufactured custom-fit earpieces with embedded electrodes. Across all participants, we are able to achieve perfect performance when we select participants' best performing tasks, mean 0\% false acceptance and 0\% false rejection rates, with data from only one earpiece with three electrodes. Furthermore, we find a 0\% false acceptance rate in a preliminary simulation of an impersonation attack. We also report some positive usability perspectives from our participants and discuss constructive directions for future work. Our results indicate that in-ear EEG could provide a discreet, convenient, and usable method for secure one-step, three-factor authentication.

@@

We find that we can achieve low false acceptance rates (FAR) and false rejection rates (FRR) with a single, 
three-electrode earpiece. Interestingly, we found that performance improves in the ear compared to a single electrode on the scalp, a site typically used for collecting EEG data.
Additionally, we find that passthoughts could not be spoofed by imposters, even when the imposters had the user's earpiece and knew their chosen passthoughts.
We discuss future work to investigate the robustness of passthoughts in day-to-day life, and raise questions about how and why passthoughts work as well as they do.

@@

Following this, we investigate custom-fit in-ear EEG technology as the platform for investigating the feasibility, performance, and usability of one-step three-factor authentication worn unobtrusively. The system we propose and test here uses the choice of a mental task or "passthought" to perform as knowledge (factor one), the uniqueness of an individual's brain activity as measured by EEG as inherence (factor two), and the physical token of custom-fit earpieces, which could easily contain a hardware key-pair, as a possession factor (factor 3). Because three-factor authentication (3FA) requires the user to submit one distinct instance of each type of authenticator, it represents the strongest level of authentication security possible. The ability to utilize all three of these security factors in a single step by performing a mental task of a few seconds is promising in pursuit of extremely strong security while maintaining a low amount of effort and obtrusiveness to the user.

@@

During the experiment, participants were seated in a comfortable position in a quiet room facing a laptop screen on which the instructions and stimuli were presented using PsychoPy \cite{peirce2007psychopy}. Each task was completed once in sets of five trials each, and then each was completed again for another five trials. Each trial was 10 seconds in length, for a total of 10 trials and 100 seconds of data collected per task. The instructions were read aloud to the participant by the experimenter, and the experiment was advanced using a pointer held in the participant's lap to minimize motion artifacts in the data. The experimenter also recorded the participant's chosen secrets for the \textit{sport}, \textit{song}, \textit{face}, \textit{speech}, and \textit{sequence} tasks and reminded the participant of these for the second set of trials.

@@

In the previous section, we trained and tested a passthoughts classifier with each task as the passthought, for all participants. Our analyses revealed the left ear alone produced decent results. Using only data from the left ear electrodes, the performance of each task for each participant is shown in Table \ref{tab:farfrrall}. For each participant we find at least one task for which they achieve 0\% FAR and FRR, indicating perfect performance for this pool of users.

\begin{table*}[t]
\centering
\begin{tabularx}{\textwidth}{lrrrrrrrrrrrrrr}
& \multicolumn{2}{c}{\textbf{P1}} & \multicolumn{2}{c}{\textbf{P2}} & \multicolumn{2}{c}{\textbf{P3}} & \multicolumn{2}{c}{\textbf{P4}} & \multicolumn{2}{c}{\textbf{P5}} & \multicolumn{2}{c}{\textbf{P6}} & \multicolumn{2}{c}{\textbf{P7}}\\
\textbf{Task} & FAR & FRR & FAR & FRR & FAR & FRR & FAR & FRR & FAR & FRR & FAR & FRR & FAR & FRR\\ \hline
Breathe & 0.0002 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0.0127 & 0 & 0 & 0 & 0\\
Breathe - open & 0.0002 & 0.0127 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0.0253 & 0 & 0 & 0 & 0\\
Face & 0.0002 & 0.0253 & 0.0005 & 0.0253 & 0 & 0.0127 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
Listen & 0 & 0 & 0.0005 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
Sequence & 0 & 0.0506 & 0.0002 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0.0007 & 0.0127\\
Song	 & 0.0002 & 0.0380 & 0 & 0 & 0	 & 0 & 0 & 0 & 0 & 0.0127 & 0 & 0 & 0.0007 & 0.0127\\
Song - open & 0.0002 & 0.0127 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0.0127\\
Speech & 0.0002 & 0 & 0.0005 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
Sport & 0.0002 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0	 & 0.0127\\
\textbf{Best Task} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\ \hline
\end{tabularx}
\caption{FAR and FRR performance of each task for each participant using data from the left ear.}
\label{tab:farfrrall}
\end{table*}

All best-performing tasks in our best-case set achieved perfect FAR and FRRs. Interestingly, \textit{breathe} appeared as the best task across all participants, except P1 whose best task was \textit{listen} and P5 who had several other best performing tasks. Given our training strategy, these results indicate that a given person's \textit{breathe} task is distinguishable not only among other tasks, but among \textit{breathe} tasks from other participants.
This task performed about as well as its open-eyes counterpart (Figure \ref{fig:meanByTask}), as did \textit{song - open} and \textit{sport} (though, interestingly, \textit{song} with eyes closed performed less well).

@@

Overall, our original training strategy achieves the lowest FAR (Table \ref{tab:compare}). Within-tasks FAR is ten times higher, and within-participants FAR is one hundred times higher, though for all scenarios still less than 1\%.
However, FRR is \textit{lower} in the within-tasks training strategy than in our original strategy's FRR. Within-participants again results in the highest FRR.

@@

While our authentication analysis establishes that passthoughts achieve low FAR and FRR when tested against other participants' passthoughts, we do not know how robust passthoughts are against a spoofing attack, in which both a participant's custom-fit electrode, and details of that participant's chosen passthought, are leaked and used by an imposter.

@@

However, since this participant's data appeared in the initial pool, the classifier may have been trained on his or her recordings as negative examples. To explore the efficacy of an outsider spoofing recordings, we repeated the same protocol with an individual who did not appear in our initial set of participants (an "outside imposter,'' Table \ref{tab:imposter}). Again, we find zero successful authentication attempts out of 1,800 attempts.

@@

In our analysis, some notable patterns emerged. First, the \textit{breathe}, \textit{breathe-open}, and \textit{listen} tasks performed exceedingly well among participants. Classifiers overall distinguished the \textit{breathe}  and \texit{listen} tasks even compared to the same tasks from other participants, implying that these tasks are expressed differently for each participant, i.e. that they have an inherence factor sufficient for authentication, even though they task do not involve an explicit secondary knowledge factor as in \textit{sport} or \texit{song}, for example. Second, we were able to achieve good results by generating feature vectors based on only 500ms (300 voltage readings across the three electrodes). This short timespan is somewhat surprising, given that some tasks (like songs) presumably rely on changes or patterns over a longer period of time.

Furthermore, likely due to the use of conductive gel, low impedances are still achievable using other participants' custom-fit earpieces, despite the uniqueness of ear canal shapes between individuals \cite{Akkermans2005}.
Nevertheless, classifiers appear to resists spoofing attacks, indicating that task-related signals are unique to individuals.
In terms of establishing a possession factor, the custom-fit earpieces could easily include a hardware keypair to sign authentication attempts in a three-factor, one-step scheme.

@@

One primary question surrounds how our passthoughts system performance will change with a greater number of users, and with more diverse data. Our system specifically trains on negative examples of incorrect users; we do not yet know how this approach will scale. At the same time, we must investigate the stability of EEG readings for a passthought over time to establish long-term usability. We must also collect EEG data from the variety of different user states: ambulatory settings, during physical exertion or exercise, under the influence of caffeine or alcohol, etc.

% We provide cursory evidence that passthoughts are difficult to spoof, though further work should expand on ours with more subjects.
Another important question surrounds how passthoughts might be cracked.
Generally, we do not understand how an individual's passthought is drawn from the distribution of EEG signals that an individual produces throughout the day. 
Given a large enough corpus of EEG data, are some passthoughts as easy to guess as \textit{password1234} is for passwords?
Future work should perform statistical analysis on passthoughts, such as clustering (perhaps with t-SNE) to better understand the space of possible passthoughts.
This work will allow us simulate cracking attempts, and to develop empirically motivated strategies for prevention, e.g., locking users out after a certain number of attempts.
This work could also reveal interesting tradeoffs between the usability or accuracy of passthoughts and their security.

Finally, our work leaves room for some clear user experience (UX) improvements.
Future work should try using dry electrodes, commonly found in consumer EEG devices, for comfort and usability. The electrodes could be grounded to the ear, instead of the forehead.
Additionally, future work could easily place a speaker inside our current custom-fit earbuds to produce working "hearables'' that can be used as ordinary headphones.

@@

As demonstrated by these preliminary results, custom-fit, in-ear EEG earpieces provide three factors of authentication in a single authentication step: thinking one's passthought, using the discreet form factor of an earpiece. In this paper, we demonstrate quite high authentication accuracy using a single sensing earpiece. By expanding our corpus of EEG readings (in population size, time, and diversity of settings), we hope to better understand the underlying distribution of EEG signals, so that we may better understand the security properties of passthoughts.
